---
title: "OCT Analysis Notes"
author: "James Edwards"
date: "5 May 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
I did some analysis of the Old County Tops 2016 results. The aim was to get some pacing schedules but I also used it as dplyr practice. It turned out to contain a lot of difficulties so I will record what I did here. My original plan was to set the file up to process other years too (they were formatted a little differently) but everything took so long I'm not going to bother.

# Import and tidying

I hid much of the processing in a separate file `OCT_PROCESS.R` which I called from the main file `OCT_analysis.R` but I'll repeat all of the relevant code here.

```{r init,  message=FALSE, warning=FALSE}
library("tidyverse")
library("readxl")
library("lubridate")
library("stringr")

setwd("E:/Dropbox/Mine/Personal/Running/Races&events/Old Counties Tops") #old laptop

excel_sheets("Results_OCT.xlsx") #gives sheetnames
num <- 1 #sheet of results to process (1 corresponds to 2016)
raw <- read_excel("Results_OCT.xlsx", sheet=num)
str(raw)
```
There were a lot of warnings because of NAs in the data frame. I was aware that there might be problems because the results were formatted with two rows per a team with some of the cells merged. What I didn't expect was that the NAs would prevent me `print`ing or viewing a small selection of rows so that I could decide what to do next. `View` and `str` still worked.

There are 20 colums. Cols 1 - 5 are: Position, team, team no, club, category. Cols 6-20 are alternating checkpoint times and placings (pairs for 6-19 then final time without placing). The positions, team no, category cols only have one entry per pair (and NAs for other runner). The times give cumulative times in the first row for each pair then leg times in the second row. Clubs may be different for each runner.

Other issues:

* The leg position columns don't have a header. 
* Some leg times are zero (missing data, not an import problem). 
* There are entires for DNF runners at the bottom of the data. I decided not to process these at all.
* The times have been imported as dates.

The last of these told me I needed to properly understand dates and times. I used the lubridate approach and converted to periods. This feels like the _correct_ way to do it but the times caused more problems later and for the specific use I wanted I could perhaps just work with decimals and convert later for output [notes on dates and times]("E:/Dropbox/Mine/Programming&IT/R/Importing data with R").

First I fixed the missing column names:
```{r variable_names}
checkp_names <- names(raw)[seq(6, 19, by=2)]
names(raw)[seq(7, 19, by=2)] <- paste0("Pos_", checkp_names)
```
For the next step I needed a new function. I also needed some similar date/time processing functions later so I give them all here (they are in `OCT_FUNC.R`. They seem quite basic time processing functions and I am suprised they do not exist already. I think I specifically needed ones that could handle a vector. The second `decimal_to_period` function rounds seconds to zero which made times easier to look at. What I really wanted was to not display seconds at all but this can't be done afaik in th eperiod class.
```{r functions, eval=FALSE}
##############
#Convert a date to a period
##############
date_to_time <- function(date){
  if(is.POSIXct(date)){
    start <- floor_date(date, unit="days")
    time <- as.period(interval(start, date))
  }else{
    time <- date
  }
  return(time)
}

##############
#Converts a decimal number to a period in HMS
##############
decimal_to_period <- function(x){
  hours <- floor(x)
  minutes <- floor((60 * x) %% 60)
  seconds <- floor((3600 * x) %% 60)
  return(period(hours=hours, minutes=minutes, seconds=seconds))
}

##############
#Converts a decimal number to a period in HM
##############
decimal_to_period <- function(x){
  hours <- floor(x)
  minutes <- round((60 * x) %% 60, 0)
  #return(period(c(hours, minutes), c("hour", "minute")))
  return(period(hours=hours, minutes=minutes))
}

##############
#Converts a decimal number to a string in H:M format
##############
decimal_to_time <- function(x){
  hours <- floor(x)
  minutes <- round((60 * x) %% 60, 0)
  return(paste0(sprintf("%01d", hours), ":", sprintf("%02d", minutes)))
}
```
Then I split the data into two frames with different purposes. The first (`main`) retained all columns but only retained the first row for each pair (cumulative times). The second (`times`) only retained names and times variables which meant that there are two rows for each pair. The `na.omit` got rid of DNFs.

```{r split_data}
source("OCT_FUNC.R")
main <- raw %>% na.omit %>% mutate_each(funs(date_to_time))
times <- raw %>% select(c(2, seq(6, 20, by=2))) %>% na.omit %>%
  mutate_each(funs(date_to_time))
```


Next I processed the `times` data frame so I just had the leg times, first storing the runner names as two vectors. NOTE: I found that using `slice` to extract rows gave incorrect output for `period` variables. No warning was given but it just took the top n rows for those variables. Subsetting with `[` was ok.
```{r times}
runners <- times$Team
n_runners <- length(runners)
runner1 <- runners[seq(1, n_runners, by=2)]
runner2 <- runners[seq(2, n_runners, by=2)]
#identical(runner1, main$Team) #should be TRUE

#Just keep leg times by keeping every second row and losing the Team variable
#times <- times %>% slice(seq(2, n_runners, by=2)) #strangely not correct
times <- times[seq(2, n_runners, by=2), ] %>% select(-1)
names(times) <- c(paste0("Leg_", checkp_names), "Leg_finish")
head(times)
```
Combining the two dataframes after adding a new column for the second team members gave the original data in tidy form. The columns needed reordering with appropriate names and I found the best way to do this was to use select with me working out the numbers manually.
```{r combine}
main <- main %>% mutate(Team2=runner2)
tidy_data <- bind_cols(main, times)
#Order columns
indices <- c(rbind(seq(6, 19, by=2), 22:28, seq(7, 19, by=2)))
tidy_data <- tidy_data %>% select(c(1, 2, 21, 3:5, indices, 20, 29))
names(tidy_data)[c(seq(7, 25, by=3), 28)] <- c(paste0("Time_", checkp_names), "Time_finish")
```

#Analysis

This was supposed to be the main part of the exercise but the data tidying took so long that I didn't have much drive left so just did the basics of what I wanted. The main step was to get leg times as a percentage of total time for each runner. 

During this I discovered that the `+` operator works for `periods` but `sum` or `rowSums` doesn't (it just adds up the seconds since that is the "numerical" part). Durations seem ok since they just record the interval in seconds and I did the division as durations.

```{r percent_leg_times}
# First need to add a Total time column to "times" df. Due to missing data I just got it from "main".
#times <- times %>% mutate(Total=rowSums(.)) #doesn't work for times
#times <- times %>% mutate(Total=Reduce(`+`, .)) #not correct as there is missing data
temp <- main %>% select(20)
times <- bind_cols(times, temp)
rm(temp)
names(times)[9] <- "Total_time"

dur_times <- times %>% mutate_each(funs(as.duration(.)) )

perc_times <- dur_times %>% mutate_each(funs(. / Total_time))
```
I didn't do much analysis on the percentage times as I found the teams' pacings fairly consistent although there is aclear trend from fastest to slowest. The code I used was a bit of a hack but would probably be more straightforward with ggplot.
```{r percent_plot, echo=F}
perc_mat <- as.matrix(perc_times)
plot(perc_mat[, 1], typ='l', ylim=c(0, 0.24))
lines(perc_mat[, 8], col=2)
legend("left", c("Leg 1", "Leg 8"), lty=1:2, col=1:2)
```

I used these percentages in two ways to get schedules. The first was based on the average of all runners and the second was using Phil and Ellie's (Phellie) percentages.

The avraging wasn't straightforward. I had to learn how to extract a vector from a tibble most methods return a tibble with one row or column. The first line demonstrates what was probably the simplest method to count the number zeroes (missing times). I want to exclude these from the average I return. I did this two ways, one by defining a function and one with an inline function. Both demonstrate the use of the pipe `.` operator.
```{r crowd_percentages}
#Average over all runners. First includes zeroes (missing entries). Second excludes zeroes.
perc_times %>% summarise_each(funs(sum(.==0))) %>% unlist(., use.names=FALSE)
mean_non_zero <- function(vec){
  return(mean(vec[vec!=0]))
}
ave_perc <- perc_times %>% summarise_each(funs(mean_non_zero)) %>% unlist(., use.names=FALSE)
ave_perc_2 <- perc_times %>% summarise_each(funs(mean(.[.!=0]))) %>% unlist(., use.names=FALSE)
identical(ave_perc, ave_perc_2)
```
I also extracted th three Lonsdale teams which didn't all use the same team name.
```{r lonsdale}
lonsdale_pos <- tidy_data %>% filter(str_detect(Club, "Lonsdale")) %>% select(1) %>% unlist(use.names = FALSE)
lonsdale <- tidy_data[lonsdale_pos, ]
perc_times[lonsdale_pos, ]
```
From this I could easily get the Phellie precentages and compare to the crowd.
```{r phellie}
phellie_perc <- perc_times[lonsdale_pos[1], ] %>% unlist(., use.names=FALSE)
round(ave_perc, 3)
round(phellie_perc,3)
```

++ Create Schedules

Producing leg and time schedules was then fairly easy. I didn't find the display of times in th `period` format compact enough so I made a function `decimal_to_time` to give a different display. Below I give the leg schedules in period format and the time schedule in my format. EDIT: I added a schedule for the top 40 teams which should be a better option.
```{r schedules}
sch_names <- c(checkp_names, "Finish", "Total")
finish_times <- seq(8, 10.5, by=0.5) #Produce schedules for these finish times

leg_schedule_crowd <- finish_times %*% t(ave_perc) %>% 
  as_data_frame %>% mutate_each(funs(decimal_to_period))
colnames(leg_schedule_crowd) <- sch_names

leg_schedule_phellie <- finish_times %*% t(phellie_perc) %>%
  as_data_frame %>% mutate_each(funs(decimal_to_period))
colnames(leg_schedule_phellie) <- sch_names

print(leg_schedule_crowd)
print(leg_schedule_phellie)

schedule_crowd <- finish_times %*% t(cumsum(ave_perc)[-9]) %>% 
  as_data_frame %>% mutate_each(funs(decimal_to_time))
colnames(schedule_crowd) <- sch_names[-9]

schedule_phellie <- finish_times %*% t(cumsum(phellie_perc)[-9]) %>% 
  as_data_frame %>% mutate_each(funs(decimal_to_time))
colnames(schedule_phellie) <- sch_names[-9]

top_crowd_perc <- perc_times[1:40, ] %>% summarise_each(funs(mean(.[.!=0]))) %>%
  unlist(., use.names=FALSE)
schedule_top_crowd <- finish_times %*% t(cumsum(top_crowd_perc)[-9]) %>% 
  as_data_frame %>% mutate_each(funs(decimal_to_time))
colnames(schedule_top_crowd) <- sch_names[-9]

print(schedule_crowd)
print(schedule_phellie, width=Inf)
print(schedule_top_crowd)
```


